%     Title page, Vitae, and Acknowledgments



%------------------------------------------------------------------
%  Title Page
%------------------------------------------------------------------

\begin{singlespace}
\begin{center}
{\huge

\vspace* {0.45in}

{\bf Efficient Excitation of Atomic\\
Rydberg States}

\vskip 0.4in

by

\vskip 0.4in

John D. Corless}

\vskip 1.0in

Submitted in Partial Fulfillment\\          
of the\\                                    
Requirements for the Degree\\               
Doctor of Philosophy                      
\vskip 0.2in

Supervised by\\                             
Professor Carlos R. Stroud, Jr.               
\vskip 1.0in
                                           
The Institute of Optics\\                   
The College\\
School of Engineering and Applied Sciences
\vskip 0.2in
                                                                                          
University of Rochester\\                   
Rochester, New York\\                       
\vskip 0.2in

1997                                      
                                                              
\end{center}
\end{singlespace}

\thispagestyle{empty}

\newpage



%------------------------------------------------------------------
%  Curriculum Vitae  
%------------------------------------------------------------------

%\centerline{\Large\bf Curriculum Vitae}
%\bigskip

\begin{singlespace}

\chapter*{Curriculum Vitae\markboth{CURRICULUM VITAE}{CURRICULUM VITAE}}
\label{cv}
\begpagestyle

\end{singlespace}

The author was born in New York, New York on November 22, 1967 and grew
up in New Jersey, Georgia and Texas.  He graduated magna cum laude with a
bachelor of science degree in physics and mathematics from Trinity University
in San Antonio, Texas.  In 1989, he began graduate studies at the Institute
of Optics.  He received the master of science degree in optical engineering
in 1992 and has carried out his doctoral research in quantum optics under the
supervision of Carlos R. Stroud, Jr.

\begin{singlespace}
\bigskip
\bigskip
\centerline{\large\bf Publications}
\bigskip
\bigskip

\setlength{\parskip}{10pt}

John D. Corless, James A. West, and C.~R. Stroud, Jr., ``Efficient and robust
population transfer with counterintuitive pulses,'' in preparation.

John D. Corless, James A. West, Jake Bromage, and C.~R. Stroud, Jr., ``Pulsed
single-mode dye laser for coherent control experiments,'' submitted to Rev. Sci.
Instrum.

John D. Corless and C.~R. Stroud, Jr., ``Optical mixing of Rydberg angular
momenta,'' submitted to Phys. Rev. Lett.

John D. Corless, James A. West, and C.~R. Stroud, Jr., ``Stochastic
fluctuations in multilevel adiabatic population transfer'' in {\em Coherence and
Quantum Optics VII}, edited by J.~H. Eberly, L. Mandel, and E. Wolf (Plenum
Press, New York, 1996).

John D. Corless, John A. Yeazell, and C.~R. Stroud, Jr., ``Shelving of
population in Rydberg wave packets by power broadening'' in {\em Proceedings of
the Sixth International Conference on Multiphoton Processes}, edited by D.~K.
Evans and S.~L. Chin (World Scientific, Singapore, 1994).

\bigskip
\end{singlespace}

\newpage

%------------------------------------------------------------------
%  Acknowledgments
%------------------------------------------------------------------

%\centerline {\Large\bf Acknowledgments}
%\bigskip

\begin{singlespace}

\chapter*{Acknowledgments\markboth{ACKNOWLEDGEMENTS}{ACKNOWLEDGEMENTS}}
\label{ack}
\begpagestyle

\end{singlespace}

It is with great pleasure that I write this most important part of my
thesis.  Many people have contributed significantly to my professional and
personal development during these enjoyable years here in Rochester, and I am
pleased to formally acknowledge them in these pages.  I must first thank
Professor Carlos R. Stroud, Jr., my thesis advisor, for his help and guidance
during the course of my graduate studies.  He has allowed me a great deal
of flexibility in choosing my research topics, which is a luxury I have
appreciated.  He always makes himself available to his graduate students,
even when he has other pressing obligations.  And finally, he fosters an
atmosphere within his group that has made the research process enjoyable
and rewarding.

There have been many people that I have interacted with here at the Institute
that deserve thanks.  My good friend Marc Adams and I shared many long nights
working together in the early years.  Professor Tom Brown was kind enough to
loan me the Nd:YAG laser used to pump the nanosecond dye lasers used in this
thesis.  The third floor research groups of Professors Stroud, Walmsley, Boyd
and Agrawal interact extensively and I have benefited from association with many
of the members of these groups, notably Andy Stentz, John Koshel, Thom Moore,
and George Fischer.  From Professor Stroud's group, I must thank former group
members Mike Noel, Zaga Gaeta, and Mark Mallalieu, with whom I have
shared many enlightening conversations.  I have had the pleasure of spending
quite a bit of time with current group members Jake Bromage and Michael Van
Leeuwen, friends and colleagues that I hold in the highest esteem.  David
Aronstein has given this thesis a most thorough reading and I thank him for
all his help.

When I first joined the group, I had the good fortune to work with John Yeazell
while he served in a post-doctoral position in Professor Stroud's group.  I have
to credit John with teaching me many things about experimental methods.  John
and I worked together on the Rydberg atom experiments reported in this thesis
and I thank him for his help with them.  I owe a great deal to my good friend
Jim West.  The two of us worked on the ``counterintuitive'' experiments
described in this thesis.  What started as a ``simple experiment in sodium''
turned into a rather extended endeavor in laser engineering and atomic
spectroscopy.  I was lucky to have Jim as a partner for this difficult
project.

My parents have always encouraged me in every aspect of my life, and this
thesis is no exception.  They always emphasized the importance of education,
and undoubtedly, these attitudes are part of the reason I decided that getting
a Ph.D. was a smart thing to do.  My wife's family has been wonderful and has
made me feel like part of the family.  I thank them for their help and
encouragement, and for making Amy.

My little daughter Sophie has only been on this earth for the past eight
weeks, yet she has enriched my life in ways that I never imagined. 
I look forward to watching her grow. Finally, it is with great love and devotion
that I acknowledge my wife Amy.  The love she has provided me has been the
foundation for my happiness, and the support she has given is the reason I have
succeeded.  I could wish for nothing greater than to spend my life with her.

This research was supported by the Corning Foundation, the Army Research Office,
and the Department of Education.

\newpage



